\chapter*{Conclusions}
\label{cha:conc}

\addcontentsline{toc}{chapter}{\nameref{cha:conc}}
\markboth{Conclusions}{Conclusions}

Mathematical modelling has become a cornerstone of the study of collective
behaviour, as the formation of macro-level structure was realised possible by
implementations of simple behavioural rules. Although model simulation can be
informative in its own right, its real power comes in comparison with
observation. Comparison of model prediction and real-world observation is an
essential part of any model-fitting and model-verification process. However,
until recently thorough comparison between collective behaviour model and data
has been limited by the availability of data of real flocking events.

In this work we acknowledged the technical challenges presented by the
recording of flocking events, and considered the affect of this on the
resulting literature. A lack of data with which to perform model-verification
has resulted in a plethora of competing models based on aprioristic assumptions
about how individuals behave and respond to stimulus. However, more recently
there have been notable attempts to bridge the gap between model and data. An
overview of this recent work, and work prior to it, was presented in
\cref{cha:lit_review}.

Backed by previous studies, we argued the importance of the quantification of
uncertainty during model-fitting. Bayesian inference, representing a
fully-probabilistic approach to parameter inference, is a natural approach to
quantifying uncertainty. The underlying philosophy of the Bayesian paradigm was
outlined in \cref{cha:bayes_intro}. Here we also introduced Markov chain Monte
Carlo methods: a class of algorithms which allow a practitioner to draw samples
from the posterior distribution, whilst only requiring knowledge of the
posterior up to a constant of proportionality.

In fitting models to data it is advisable to first consider simple models, and
to only increase model-complexity as necessary. The Vicsek model in
\cref{cha:model_dev} was introduced as one such simple model, accounting for
directional changes as some combination of alignment interactions and noise. We
questioned the biological-realism the zonal interaction rule implemented by
Vicsek represents, instead advocating the merits of continuously differentiable
interaction rules. As well as biological-realism, we acknowledged the
difficulties which discontinuous interaction rules pose for parameter
inference.

Despite understandings of the importance of biological variation in nature, the
majority of models in the literature assume that all agents behave identically
when presented with the same stimulus. With this shortcoming in mind we
introduced models with hierarchical structure: allowing intra-flock behavioural
and biological variation.

\cref{cha:sim_studies} was a chapter of simulation studies. Simulation studies
are a useful tool in assessing the viability of performing parameter inference
on real data. The models introduced and developed in the preceding chapter
were forward simulated for known parameter values. We then demonstrated that
the Metropolis--Hastings and NUTS algorithms could be used to accurately infer
the true parameter values from observations of the simulated flocks alone.

Having demonstrated the efficacy of our approach to inference on simulated
data, we proceeded to re-use our statistical machinery to perform inference on
real data. In \cref{cha:sheep} we introduced a dataset consisting of three
recording events of flocking sheep. To this data we fit variations of the
Vicsek model as introduced in \cref{cha:model_dev}. The within-sample
predictive performance of these models was quantified with information
criteria. Models were then ranked by their performance. We observed that
continuous interaction rules consistently outranked their discrete counterpart,
as well as observing evidence of non-normally distributed noise. In general the
models provided a good fit to data, however in one of the sequences there was
evidence of the existence of behaviours \emph{not} accounted for in our models.
We speculated that these behaviours could be the result of some unaccounted for
interactions with the external environment, or additional inter-individual
interactions such as attraction or repulsion.

As it is impossible to predict when and where a given flocking event may occur
next, there can become a frustrating ``right-place-right-time'' component to
data collection. Flocking events may be captured by establishing fixed-location
recording equipment where a researcher believes a flocking event may occur in
the future. However, the stationary nature of the recording equipment can
result in capturing incomplete flocking events. To evaluate the likelihood of
the data---a critical component of any model-fitting, Bayesian or
otherwise---the observation of every individual in the flock, at every recorded
frame, is required. A naive solution to this problem is to discard every frame
in which the entire flock isn't visible. However, this approach has the
potential to \emph{drastically} reduce the amount of data available for
analysis. In \cref{cha:missing} we presented an alternative approach which
incorporates the additional uncertainty introduced by missingness into our
posterior distributions. Simulation studies were performed which showed that
our approach consistently outperformed the naive approach of discarding data.

The methodology we developed in \cref{cha:missing} was used on a real dataset
exhibiting missingness. This dataset detailed the movements of ducks
foraging on the surface of a lake, and was captured with fixed-location
recording equipment. A large number of data points were missing from this
sequence, as much movement took place out-of-frame. This presented a
computationally demanding, high-dimensional inference problem. In fitting a
variation of Vicsek to this data, we observed evidence that there is no direct
alignment interaction between individuals. We postulated the existence of
repulsion or attraction behaviours, not accounted for by our model, which
maintain the flocks highly polarised structure. Although our model could not
explain the interactions between individuals, it was only by accounting for the
missing observations that we were able to reach this conclusion.

Throughout this thesis we demonstrated a fully-Bayesian approach to fitting
models of collective behaviour to real and simulated data. We advocated
consideration of simple models over the needlessly complex, and the addition of
complexity only when simpler models perform poorly. Although this work only
considered a small number of possible candidate models, the framework we
outlined can be used to compare any number of competing models.

We demonstrated to the reader that although fitting models of collective motion
to real data is difficult, the process can be \emph{greatly} eased by using
existing, general purpose inference tools such as the Stan programming
language. Such software will undoubtedly outperform the hand-coded MCMC
algorithms of all but the most talented computer programmers. To the reader we
also emphasise the utility of simulation studies. Simulation studies represent
an invaluable opportunity to assess the feasibility of parameter inference on
any model of interest, and give scope to develop \emph{and} debug inference
code in an environment where the answer is known. Although methodology was
outlined which allowed the researcher to account for missing data within
recordings of flocking events, the computational cost that this approach incurs
is so large that in the first instance the greatest effort should be made to
avoid missing data during the recording process.
