\graphicspath{{fig/sheep/}}

\chapter{Case study I: flocking sheep}
\label{cha:sheep}

In \cref{cha:sim_studies} we showed that it is possible to infer the parameters
of a number of agent based models from simulated data. Performing simulation
studies before attempting inference on real data is recommended as it allows an
opportunity to troubleshoot and assess the accuracy of our inference schemes.
The schemes constructed to perform inference on simulated data can now be
reused with little to no alteration, to perform inference on real data. With
the confidence instilled by the success of our simulation studies, if we find
our schemes unable to fit a model to real data, we can be confident that this
is because of a discrepancy between the model and data, rather than an error in
our fitting process.

This chapter shall focus on data of flocking sheep, unseen before in the
literature. Aerial footage of flocking sheep was recorded by a commercially
available drone, and the position of sheep in every frame was extracted using
custom tracking software. In total, three flocking sequences were extracted and
analysed. Global models considered in \cref{sec:global_models} will be fit to
these sequences. The resulting fits will be assessed for validity and ranked
for performance by Akaike Information Criteria.

\section{Data}
\label{sec:data}

Aerial footage of flocking sheep was captured with a commercially available DJI
Phantom $3$ drone equipped with a built in high-definition camera, representing
a frame resolution of $4000\times3000$ pixels. The aerial footage was recorded
at a rate of $24$ frames per second. In total, three distinct flocking events
were captured and analysed. Each flocking event consisted of the same $45$
sheep.

Captured flocking events took place in large open spaces, in a natural
environment which the sheep were already familiar with. In some cases the
flocking events were initiated by the movements of a quad bike, but were then
allowed to develop naturally and unimpeded. Events generally took place away
from hard boundaries, such as fences and trees, but the influence of these
objects from a distance cannot be ruled-out.



